---
title: "Atlas neighbors"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.lazy = FALSE,
  cache.path = "cache/11-atlas-neighbors.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(batchelor)
library(BiocNeighbors)
library(MouseGastrulationData)
library(scater)
library(scran)
```

Load CRAN packages:

```{r}
library(ggforce)
library(pheatmap)
```

Source user functions:

```{r}
source("code/splitByCol.R")
source("code/weightedkNN.R")
```

## Neighbors

Read experiment object:

```{r}
sce <- readRDS("output/10-atlas-integration.rds")
```

Define source annotation:

```{r}
sce$source <- ifelse(sce$batch == "LPMP_EHF", "query", "reference")
```

Split experiment object by source:

```{r}
sce <- splitByCol(sce, sce$source)
```

Extract corrected PCA matrix for reference and query data:

```{r}
dim <- list(
  reference = reducedDim(sce$reference, "corrected"),
  query = reducedDim(sce$query, "corrected")
)
```

Find kNN in reference data for each point in query data:

```{r}
set.seed(0011110001)

knn <- queryKNN(X = dim$reference, query = dim$query, k = 20) # k = 20 to match fastMNN
```

Create matrix of kNN annotations:

```{r}
knn$cell <- structure(sce$reference$cell[knn$index], .Dim = dim(knn$index))

knn$stage <- structure(sce$reference$stage[knn$index], .Dim = dim(knn$index))

knn$celltype <- structure(sce$reference$celltype[knn$index], .Dim = dim(knn$index))
```

Plot UMAP embedding of kNN cells:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(sce$reference, use.coldata = c("stage", "celltype"), use.dimred = "UMAP"),
  query = makePerCellDF(sce$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[sce$reference$cell %in% knn$cell] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN stages:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(sce$reference, use.coldata = c("stage", "celltype"), use.dimred = "UMAP"),
  query = makePerCellDF(sce$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[sce$reference$stage %in% knn$stage] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN celltypes:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(sce$reference, use.coldata = c("stage", "celltype"), use.dimred = "UMAP"),
  query = makePerCellDF(sce$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[dat$reference$celltype %in% knn$celltype] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage, drop = FALSE) +  
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Annotate query data with weighted kNN annotations:

```{r}
sce$query$cell.mapped <- as.list(as.data.frame(t(knn$cell)))

sce$query$stage.mapped <- weightedkNN(knn, class = "stage")

sce$query$celltype.mapped <- weightedkNN(knn, class = "celltype")
```

Self-annotate reference data:

```{r}
sce$reference$cell.mapped <- sce$reference$cell

sce$reference$stage.mapped <- sce$reference$stage

sce$reference$celltype.mapped <- sce$reference$celltype
```

Subset reference data to kNN annotations:

```{r}
id1 <- sce$reference$celltype %in% sce$query$celltype.mapped

id2 <- sce$reference$stage %in% sce$query$stage.mapped

idx <- id1 & id2

sce$reference <- sce$reference[ , idx]
```

Combine reference and query data:

```{r}
sce <- cbind(sce$reference, sce$query)
```

## Integration

### Variance modelling

Model the per-gene variance:

```{r}
dec <- modelGeneVar(sce, block = sce$sample)
```

### Feature selection

Read blacklist genes and subset by category:

```{r}
nan <- read.csv("data/blacklist.csv")

nan <- subset(nan, CATEGORY %in% c("CC", "MT", "RP", "XIST", "Y"))
```

Identify blacklist genes to ignore in feature selection:

```{r}
dec$blacklist <- FALSE

dec$blacklist[rownames(dec) %in% nan$ENSEMBL] <- TRUE
```

Identify highly variable genes:

```{r}
hvg <- subset(dec, blacklist == FALSE)

hvg <- getTopHVGs(hvg, var.field = "bio", var.threshold = 0, fdr.field = "p.value", fdr.threshold = 0.05)
```

### Batch normalization

Perform scaling normalization:

```{r}
sce <- multiBatchNorm(sce, batch = sce$sample)
```

### Batch integration

Create sequential merge order:

```{r}
dat <- as.data.frame(table(sce$sample, sce$stage), stringsAsFactors = FALSE)

colnames(dat) <- c("sample", "stage", "ncells")

dat <- subset(dat, ncells > 0)

dat <- subset(dat, sample %in% AtlasSampleMetadata$sample)

lvl <- c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5")

dat$stage <- factor(dat$stage, levels = lvl)

dat <- dat[order(dat$stage, dat$ncells, decreasing = TRUE), ]

idx <- c(dat$sample, "LPMP_EHF")
```

Correct experiments using nearest neighbours:

```{r}
set.seed(0000110010)

mnn <- correctExperiments(
  sce,
  batch = sce$sample,
  subset.row = hvg,
  correct.all = TRUE,
  PARAM = FastMnnParam(merge.order = idx)
)
```

Store combined variance model:

```{r}
metadata(mnn)$modelGeneVar <- dec
```

Store highly variable genes:

```{r}
metadata(mnn)$getTopHVGs <- hvg
```

## Evaluation

Perform UMAP on the corrected PCA matrix:

```{r}
set.seed(42)

mnn <- runUMAP(mnn, dimred = "corrected", n_neighbors = 20, min_dist = 0.7) # same settings as reference analysis
```

Split experiment object by source:

```{r}
mnn <- splitByCol(mnn, mnn$source)
```

Plot UMAP embedding of mapped stage and mapped celltype:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped", "source"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = c("stage.mapped", "celltype.mapped", "source"), use.dimred = "UMAP")
)

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2, colour = celltype.mapped), size = 2, shape = 4, inherit.aes = FALSE, show.legend = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_wrap(~ stage.mapped) +  
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of mapped stage and marker-expression celltype:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped", "source"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = c("stage.mapped", "celltype", "source"), use.dimred = "UMAP")
)

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2, colour = celltype), size = 2, shape = 4, inherit.aes = FALSE, show.legend = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_wrap(~ stage.mapped) +  
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

## Re-Annotation

Extract corrected PCA matrix for reference and query data:

```{r}
dim <- list(
  reference = reducedDim(mnn$reference, "corrected"),
  query = reducedDim(mnn$query, "corrected")
)
```

Find kNN in reference data for each point in query data:

```{r}
set.seed(0011110001)

knn <- queryKNN(X = dim$reference, query = dim$query, k = 20) # k = 20 to match fastMNN
```

Create matrix of kNN annotations:

```{r}
knn$cell <- structure(mnn$reference$cell[knn$index], .Dim = dim(knn$index))

knn$stage <- structure(mnn$reference$stage[knn$index], .Dim = dim(knn$index))

knn$celltype <- structure(mnn$reference$celltype[knn$index], .Dim = dim(knn$index))
```

Plot UMAP embedding of kNN cells:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[mnn$reference$cell %in% knn$cell] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage.mapped, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN stages:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[mnn$reference$stage %in% knn$stage] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage.mapped, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN celltypes:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[dat$reference$celltype %in% knn$celltype] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage.mapped, drop = FALSE) +  
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

## Summary

### Output

### Session

Print session information:

```{r}
sessionInfo()
```
