---
title: "Atlas neighbors"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.lazy = FALSE,
  cache.path = "cache/11-atlas-neighbors.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(BiocNeighbors)
library(MouseGastrulationData)
library(batchelor)
library(scater)
library(scran)
library(scuttle)
```

Load CRAN packages:

```{r}
library(ggforce)
library(pheatmap)
```

Source user functions:

```{r}
source("code/splitByCol.R")
source("code/weightedKNN.R")
```

## Neighbors

Read experiment object:

```{r}
sce <- readRDS("output/10-atlas-integration.rds")
```

Define source annotation:

```{r}
sce$source <- ifelse(sce$batch == "LPMP_EHF", "query", "reference")
```

Split experiment object by source:

```{r}
sce <- splitByCol(sce, sce$source)
```

Extract corrected PCA matrix for reference and query data:

```{r}
dim <- list(
  reference = reducedDim(sce$reference, "corrected"),
  query = reducedDim(sce$query, "corrected")
)
```

Find kNN in reference data for each point in query data:

```{r}
set.seed(0011110001)

knn <- queryKNN(X = dim$reference, query = dim$query, k = 20) # k = 20 to match fastMNN
```

Create matrix of kNN annotations:

```{r}
knn$cell <- structure(sce$reference$cell[knn$index], .Dim = dim(knn$index))

knn$stage <- structure(sce$reference$stage[knn$index], .Dim = dim(knn$index))

knn$celltype <- structure(sce$reference$celltype[knn$index], .Dim = dim(knn$index))
```

Plot UMAP embedding of kNN cells:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(sce$reference, use.coldata = c("stage", "celltype"), use.dimred = "UMAP"),
  query = makePerCellDF(sce$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[sce$reference$cell %in% knn$cell] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN stages:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(sce$reference, use.coldata = c("stage", "celltype"), use.dimred = "UMAP"),
  query = makePerCellDF(sce$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[sce$reference$stage %in% knn$stage] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN celltypes:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(sce$reference, use.coldata = c("stage", "celltype"), use.dimred = "UMAP"),
  query = makePerCellDF(sce$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[dat$reference$celltype %in% knn$celltype] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage, drop = FALSE) +  
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Annotate query data with weighted kNN annotations:

```{r}
sce$query$cell.mapped <- as.list(as.data.frame(t(knn$cell)))

sce$query$stage.mapped <- weightedKNN(knn, class = "stage")

sce$query$celltype.mapped <- weightedKNN(knn, class = "celltype")
```

Self-annotate reference data:

```{r}
sce$reference$cell.mapped <- sce$reference$cell

sce$reference$stage.mapped <- sce$reference$stage

sce$reference$celltype.mapped <- sce$reference$celltype
```

Subset reference data to kNN annotations:

```{r}
id1 <- sce$reference$celltype %in% sce$query$celltype.mapped

id2 <- sce$reference$stage %in% sce$query$stage.mapped

idx <- id1 & id2

sce$reference <- sce$reference[, idx]
```

Combine reference and query data:

```{r}
sce <- cbind(sce$reference, sce$query)
```

## Integration

### Variance modelling

Model the per-gene variance:

```{r}
dec <- modelGeneVar(sce, block = sce$sample)
```

### Feature selection

Read blacklist genes and subset by category:

```{r}
nan <- read.csv("data/blacklist.csv")

nan <- subset(nan, CATEGORY %in% c("CC", "MT", "RP", "XIST", "Y"))
```

Ignore blacklist genes in feature selection:

```{r}
dec$blacklist <- FALSE

dec$blacklist[rownames(dec) %in% nan$ENSEMBL] <- TRUE
```

Identify highly variable genes:

```{r}
hvg <- subset(dec, blacklist == FALSE)

hvg <- getTopHVGs(hvg, var.field = "bio", var.threshold = 0, fdr.field = "p.value", fdr.threshold = 0.05)
```

### Batch normalization

Perform scaling normalization:

```{r}
sce <- multiBatchNorm(sce, batch = sce$sample)
```

### Batch integration

Create sequential merge order:

```{r}
tab <- as.data.frame(table(sce$sample, sce$stage), stringsAsFactors = FALSE)

colnames(tab) <- c("sample", "stage", "ncells")

tab <- subset(tab, ncells > 0)

tab <- subset(tab, sample %in% AtlasSampleMetadata$sample)

lvl <- c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5")

tab$stage <- factor(tab$stage, levels = lvl)

tab <- tab[order(tab$stage, tab$ncells, decreasing = TRUE), ]

idx <- c(tab$sample, "LPMP_EHF")
```

Correct experiments using nearest neighbours:

```{r}
set.seed(0000110010)

mnn <- correctExperiments(
  sce,
  batch = sce$sample,
  subset.row = hvg,
  correct.all = TRUE,
  PARAM = FastMnnParam(merge.order = idx)
)
```

Store combined variance model:

```{r}
metadata(mnn)$modelGeneVar <- dec
```

Store highly variable genes:

```{r}
metadata(mnn)$getTopHVGs <- hvg
```

## Diagnostics

Check proportion of lost variance:

```{r}
var <- metadata(mnn)$merge.info$lost.var

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Reds"))(100)

brk <- seq(0, max(var), length.out = 101)

pheatmap(var[, idx], color = col, display_numbers = TRUE, number_format = "%.2f", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Cluster on the corrected PCA to obtain a partitioning of the cells:

```{r}
snn <- buildSNNGraph(mnn, type = "jaccard", use.dimred = "corrected")

com <- igraph::cluster_louvain(snn)
```

Tabulate number of cells from multiple batches within each cluster:

```{r}
tab <- table(Cluster = com$membership, Batch = mnn$batch)

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Blues"))(100)

brk <- seq(0, max(tab), length.out = 101)

pheatmap(tab[, idx], color = col, display_numbers = TRUE, number_format = "%d", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Perform TSNE on the corrected PCA matrix:

```{r}
set.seed(42)

mnn <- runTSNE(mnn, dimred = "corrected", perplexity = 120) # same settings as reference analysis
```

Visualize the corrected PCA using a TSNE plot:

```{r fig.width = 9, fig.height = 9}
ggcells(mnn, aes(TSNE.1, TSNE.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1, show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours, na.value = "#000000") + 
  labs(x = "TSNE 1", y = "TSNE 2") + 
  facet_wrap(~ sample) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

Perform UMAP on the corrected PCA matrix:

```{r}
set.seed(42)

mnn <- runUMAP(mnn, dimred = "corrected", n_neighbors = 20, min_dist = 0.7) # same settings as reference analysis
```

Visualize the corrected PCA using a UMAP plot:

```{r fig.width = 9, fig.height = 9}
ggcells(mnn, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1, show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours, na.value = "#000000") + 
  labs(x = "UMAP 1", y = "UMAP 2") + 
  facet_wrap(~ sample) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

Plot TSNE embedding of mapped stage and mapped celltype:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  all = makePerCellDF(mnn, use.coldata = NULL, use.dimred = "TSNE"),
  col = makePerCellDF(mnn, use.coldata = c("stage.mapped", "celltype.mapped", "source"), use.dimred = "TSNE")
)

dat$col$source <- factor(dat$col$source, levels = c("query", "reference"), labels = c("LPMP EHF", "Embryo Atlas"))

ggplot() + 
  geom_point(data = dat$all, aes(TSNE.1, TSNE.2), colour = "grey92", size = 0.1) + 
  geom_point(data = dat$col, aes(TSNE.1, TSNE.2, colour = celltype.mapped), size = 0.1) + 
  scale_colour_manual(name = "Celltype (Mapped)", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(source ~ stage.mapped) +  
  labs(subtitle = "Stage (Mapped)") + 
  theme_no_axes() + 
  theme(aspect.ratio = 1, plot.subtitle = element_text(hjust = 0.5))
```

Plot UMAP embedding of mapped stage and mapped celltype:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  all = makePerCellDF(mnn, use.coldata = NULL, use.dimred = "UMAP"),
  col = makePerCellDF(mnn, use.coldata = c("stage.mapped", "celltype.mapped", "source"), use.dimred = "UMAP")
)

dat$col$source <- factor(dat$col$source, levels = c("query", "reference"), labels = c("LPMP EHF", "Embryo Atlas"))

ggplot() + 
  geom_point(data = dat$all, aes(UMAP.1, UMAP.2), colour = "grey92", size = 0.1) + 
  geom_point(data = dat$col, aes(UMAP.1, UMAP.2, colour = celltype.mapped), size = 0.1) + 
  scale_colour_manual(name = "Celltype (Mapped)", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(source ~ stage.mapped) + 
  labs(subtitle = "Stage (Mapped)") + 
  theme_no_axes() + 
  theme(aspect.ratio = 1, plot.subtitle = element_text(hjust = 0.5))
```

## Annotation

Split experiment object by source:

```{r}
mnn <- splitByCol(mnn, mnn$source)
```

Extract corrected PCA matrix for reference and query data:

```{r}
dim <- list(
  reference = reducedDim(mnn$reference, "corrected"),
  query = reducedDim(mnn$query, "corrected")
)
```

Find kNN in reference data for each point in query data:

```{r}
set.seed(0011110001)

knn <- queryKNN(X = dim$reference, query = dim$query, k = 20) # k = 20 to match fastMNN
```

Create matrix of kNN annotations:

```{r}
knn$cell <- structure(mnn$reference$cell[knn$index], .Dim = dim(knn$index))

knn$stage <- structure(mnn$reference$stage[knn$index], .Dim = dim(knn$index))

knn$celltype <- structure(mnn$reference$celltype[knn$index], .Dim = dim(knn$index))
```

Plot UMAP embedding of kNN cells:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[mnn$reference$cell %in% knn$cell] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage.mapped, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN stages:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[mnn$reference$stage %in% knn$stage] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage.mapped, drop = FALSE) + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot UMAP embedding of kNN celltypes:

```{r fig.width = 12, fig.height = 9}
dat <- list(
  reference = makePerCellDF(mnn$reference, use.coldata = c("stage.mapped", "celltype.mapped"), use.dimred = "UMAP"),
  query = makePerCellDF(mnn$query, use.coldata = NULL, use.dimred = "UMAP")
)

dat$reference$neighbor <- "Non-neighbor"

dat$reference$neighbor[dat$reference$celltype %in% knn$celltype] <- "Neighbor"

dat$reference$neighbor <- factor(dat$reference$neighbor, levels = c("Neighbor", "Non-neighbor"))

ggplot(dat$reference, aes(UMAP.1, UMAP.2, colour = celltype.mapped)) + 
  geom_point(size = 0.1) + 
  geom_point(data = dat$query, aes(UMAP.1, UMAP.2), size = 1, shape = 4, inherit.aes = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1))) + 
  facet_grid(neighbor ~ stage.mapped, drop = FALSE) +  
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Annotate query data with weighted kNN annotations:

```{r}
mnn$query$cell.remapped <- as.list(as.data.frame(t(knn$cell)))

mnn$query$stage.remapped <- weightedKNN(knn, class = "stage")

mnn$query$celltype.remapped <- weightedKNN(knn, class = "celltype")
```

Self-annotate reference data:

```{r}
mnn$reference$cell.remapped <- mnn$reference$cell

mnn$reference$stage.remapped <- mnn$reference$stage

mnn$reference$celltype.remapped <- mnn$reference$celltype
```

Compare mapped and remapped stage annotations:

```{r}
tab <- table(Mapped = mnn$query$stage.mapped, Remapped = mnn$query$stage.remapped)

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Blues"))(100)

brk <- seq(0, max(tab), length.out = 101)

pheatmap(tab, color = col, display_numbers = TRUE, number_format = "%d", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Compare mapped and remapped celltype annotations:

```{r}
tab <- table(Mapped = mnn$query$celltype.mapped, Remapped = mnn$query$celltype.remapped)

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Blues"))(100)

brk <- seq(0, max(tab), length.out = 101)

pheatmap(tab, color = col, display_numbers = TRUE, number_format = "%d", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Combine reference and query data:

```{r}
mnn <- cbind(mnn$reference, mnn$query)
```

## Summary

### Output

Save experiment object to disk:

```{r}
saveRDS(mnn, file = "output/11-atlas-neighbors.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
