---
title: "Atlas trajectory"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/12-atlas-trajectory.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(scater)
library(scran)
library(scuttle)
library(slingshot)
```

Load CRAN packages:

```{r}
library(ggforce)
```

Define colour palettes:

```{r}
EmbryoStageColours <- c(
  "E6.5"  = "#D53E4F",
  "E6.75" = "#F46D43",
  "E7.0"  = "#FDAE61",
  "E7.25" = "#FEE08B",
  "E7.5"  = "#FFFFBF",
  "E7.75" = "#E6F598",
  "E8.0"  = "#ABDDA4",
  "E8.25" = "#66C2A5",
  "E8.5"  = "#3288BD",
  "mixed_gastrulation" = "#A9A9A9"
)

EmbryoCelltypeColours <- MouseGastrulationData::EmbryoCelltypeColours

EmbryoSourceColours <- c("query" = "#E15759", "reference" = "#BAB0AC")
```

Define column data labels:

```{r}
EmbryoSourceLabels <- c("reference" = "Mouse Embryo Atlas", "query" = "Mouse LPMP EHF")
```

Read experiment object:

```{r}
sce <- readRDS("output/11-atlas-neighbors.rds")
```

Make feature names unique:

```{r}
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$gene_id, rowData(sce)$gene_name)
```

## Trajectory

Extract query and kNN cells:

```{r}
id1 <- sce$cell[sce$source == "query"]

id2 <- unique(sort(unlist(sce$cell.remapped[sce$source == "query"])))

ind <- sce$cell %in% c(id1, id2)

sce <- sce[, ind]
```

Source function to calculate distance with MNN coordinates:

```{r}
source("code/dist_clusters_diag.R")
```

Perform lineage inference with Slingshot:

```{r}
set.seed(378891116)

sce <- slingshot(
  data = sce,
  clusterLabels = sce$celltype.remapped,
  reducedDim = "corrected",
  dist.fun = dist_clusters_diag,
  approx_points = 100
)
```

Plot principal curves on TSNE embedding:

```{r}
cur <- embedCurves(sce, "TSNE")

cur <- slingCurves(cur)

cur <- lapply(cur, function(x) {
  
  d <- x$s[x$ord, ]
  
  d <- as.data.frame(d)
  
  rownames(d) <- NULL
  
  colnames(d) <- c("TSNE.1", "TSNE.2")
  
  return(d)

})

cur <- dplyr::bind_rows(cur, .id = "curve")

dat <- makePerCellDF(sce, use.coldata = c("source", "celltype.remapped"), use.dimred = "TSNE")

dat <- split(dat, dat$source)

ggplot() + 
  geom_point(data = dat$reference, mapping = aes(TSNE.1, TSNE.2, fill = celltype.remapped), colour = "transparent", shape = 21, inherit.aes = FALSE) + 
  geom_point(data = dat$query, mapping = aes(TSNE.1, TSNE.2), shape = 1, inherit.aes = FALSE) + 
  geom_path(data = cur, mapping = aes(TSNE.1, TSNE.2, colour = curve)) + 
  scale_fill_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  scale_colour_brewer(name = "Curve", palette = "Set1") + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Plot principal curves on UMAP embedding:

```{r}
cur <- embedCurves(sce, "UMAP")

cur <- slingCurves(cur)

cur <- lapply(cur, function(x) {
  
  d <- x$s[x$ord, ]
  
  d <- as.data.frame(d)
  
  rownames(d) <- NULL
  
  colnames(d) <- c("UMAP.1", "UMAP.2")
  
  return(d)

})

cur <- dplyr::bind_rows(cur, .id = "curve")

dat <- makePerCellDF(sce, use.coldata = c("source", "celltype.remapped"), use.dimred = "UMAP")

dat <- split(dat, dat$source)

ggplot() + 
  geom_point(data = dat$reference, mapping = aes(UMAP.1, UMAP.2, fill = celltype.remapped), colour = "transparent", shape = 21, inherit.aes = FALSE) + 
  geom_point(data = dat$query, mapping = aes(UMAP.1, UMAP.2), shape = 1, inherit.aes = FALSE) + 
  geom_path(data = cur, mapping = aes(UMAP.1, UMAP.2, colour = curve)) + 
  scale_fill_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  scale_colour_brewer(name = "Curve", palette = "Set1") + 
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

## Differential


```{r}
vec <- split(sce$slingPseudotime_4, sce$batch)

vec <- lapply(vec, unique)

vec <- sapply(vec, function(x) length(x) > 6)

ids <- names(vec)[vec]

ind <- sce$batch %in% ids

res <- TSCAN::testPseudotime(
  x = sce[, ind],
  pseudotime = sce$slingPseudotime_4[ind],
  block = sce$batch[ind],
  assay.type = "logcounts"
)

sig <- subset(res, FDR < 0.05)

sig <- sig[order(abs(sig$logFC), decreasing = TRUE),  ]
```

```{r}
plotHeatmap(
  object = sce[, !is.na(sce$slingPseudotime_4)],
  order_columns_by = "slingPseudotime_4", 
  colour_columns_by = c("source", "celltype.remapped"),
  features = rownames(head(sig, n = 50)),
  center = TRUE
)
```



## Differential

Fit the NB-GAM model with batch effect design matrix on highly variable genes:

```{r}
set.seed(430396281)

mod <- model.matrix(~ 0 + source, colData(sce))

colnames(mod) <- c("reference", "test")

idx <- which(rowData(sce)$gene_id %in% metadata(sce)$getTopHVGs)

fit <- fitGAM(counts = counts(sce), sds = SlingshotDataSet(sce), U = mod, genes = idx)
```

Test whether average gene expression is associated with pseudotime:

```{r}
res <- associationTest(fit)

res$FDR <- p.adjust(res$pvalue, method = "fdr")

res$symbol <- rowData(sce)$gene_name[match(rownames(res), rowData(sce)$gene_id)]

res <- subset(res, FDR < 0.05)

res <- res[order(res$pvalue), ]
```

Create table of associationTest results:

```{r}
knitr::knit_print(DT::datatable(res))
```

Function to plot expression of genes associated with pseudotime:

```{r}
plotPseudotime <- function(object, features) {
  
  colnames(object) <- paste("cell", seq_len(ncol(object)), sep = "-")
  
  object <- object[features, order(object$slingPseudotime_1)]
  
  x <- assay(object, "reconstructed")
  
  rownames(x) <- rowData(object)$gene_name

  M <- rowMeans(x, na.rm = TRUE)
  
  DF <- ncol(x) - 1L
  
  IsNA <- is.na(x)
  
  if (any(IsNA)) {
    
    mode(IsNA) <- "integer"
    
    DF <-  DF - rowSums(IsNA)
    
    DF[DF == 0L] <- 1L
  
  }

  x <- x - M
  
  V <- rowSums(x^2L, na.rm = TRUE) / DF
	
  x <- x / sqrt(V + 0.01)
  
  pheatmap.cluster_rows <- hclust(dist(x, method = "euclidean"), method = "complete")

  pheatmap.color <- colorRampPalette(rev(brewer.pal(n = 5, name = "RdBu")))(100)
  
  l <- max(abs(x))
  
  pheatmap.breaks <- seq(-l, l, length.out = 101)
  
  pheatmap.annotation_col <- data.frame(
    Celltype = object$celltype, 
    Stage = object$stage,
    Source = object$source,
    row.names = colnames(object))
  
  pheatmap.annotation_colors = list(
    Celltype = EmbryoCelltypeColours[names(EmbryoCelltypeColours) %in% object$celltype],
    Stage = setNames(tableau_color_pal("Tableau 10")(5), sort(unique(sce$stage))),
    Source = setNames(tableau_color_pal("Classic Green-Orange 6")(2), sort(unique(sce$source)))
  )

  pheatmap(
    mat = x,
    color = pheatmap.color,
    breaks = pheatmap.breaks,
    cluster_rows = pheatmap.cluster_rows,
    cluster_cols = FALSE,
    annotation_col = pheatmap.annotation_col,
    annotation_colors = pheatmap.annotation_colors,
    show_colnames = FALSE
  )

}
```

Plot expression of genes associated with pseudotime:

```{r fig.width = 12, fig.height = 8}
plotPseudotime(sce, head(rownames(res), n = 50))
```

## Summary

### Output

Save integrated experiment object to disk:

```{r}
saveRDS(sce, file = "output/12-atlas-trajectory.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
