---
title: "CellChat"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/cellchat.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

```{r, include = FALSE}
DT::datatable(matrix()) # Import datatable options, quietly
```

Load Bioconductor packages:

```{r}
library(scuttle)
```

Load CRAN packages:

```{r}
library(hues)
library(circlize)
```

Load Github packages:

```{r}
library(CellChat)
```

## Data

Read experiment object:

```{r}
sce <- readRDS("output/secondary-analysis.rds")
```

Make feature names unique:

```{r}
rownames(sce) <- uniquifyFeatureNames(
  ID = rowData(sce)$gene_id,
  names = rowData(sce)$gene_name
)
```

## CellChat

Create a CellChat object from a SingleCellExperiment object:

```{r}
cellchat <- createCellChat(object = sce, group.by = "label")
```

Set the ligand-receptor interaction database:

```{r}
cellchat@DB <- CellChatDB.mouse
```

Subset the expression data of signaling genes for saving computation cost:

```{r}
cellchat <- subsetData(cellchat)
```

Identify over-expressed signaling genes associated with each cell group:

```{r}
cellchat <- identifyOverExpressedGenes(cellchat)
```

Identify over-expressed ligand-receptor interactions (pairs) within the used CellChatDB:

```{r}
cellchat <- identifyOverExpressedInteractions(cellchat)
```

Project gene expression data onto a protein-protein interaction network:

```{r}
cellchat <- projectData(cellchat, PPI.mouse)
```

Compute the communication probability/strength between any interacting cell groups:

```{r}
cellchat <- computeCommunProb(cellchat)
```

Compute the communication probability on signaling pathway level by summarizing all related ligands/receptors:

```{r}
cellchat <- computeCommunProbPathway(cellchat)
```

Calculate the aggregated network by counting the number of links or summarizing the communication probability:

```{r}
cellchat <- aggregateNet(cellchat)
```

Returns a data frame consisting of all the inferred cell-cell communications:

```{r}
net <- subsetCommunication(cellchat)
```

Print interactive table of all the inferred cell-cell communications:

```{r}
cat(knitr::knit_print(DT::datatable(dat)))
```

<!-- ## Results {.tabset} -->

<!-- Define chord diagram function: -->

<!-- ```{r} -->
<!-- chordDiagram <- function(x, title = NULL) { -->

<!--   ## Create data frame with connections -->

<!--   dat <- data.frame( -->
<!--     from = paste(x$source, x$ligand, sep = "."),  -->
<!--     to = paste(x$target, x$receptor, sep = ".") -->
<!--   ) -->

<!--   ## Create connections annotation -->

<!--   ann <- setNames( -->
<!--     object = c(x$ligand, x$receptor),  -->
<!--     nm = c(x$source, x$target) -->
<!--   ) -->

<!--   ## Filter duplicated connections -->

<!--   ids <- paste(names(ann), ann, sep = ".") -->

<!--   ann <- ann[!duplicated(ids)] -->


<!--   ## Order connections by name and value -->

<!--   ids <- paste(names(ann), ann, sep = ".") -->

<!--   ann <- ann[order(ids)] -->


<!--   ## Create LR colour palette -->

<!--   ids <- unique(ann) -->

<!--   col <- iwanthue(n = length(ids)) -->

<!--   names(col) <- ids -->

<!--   col <- col[ann] -->

<!--   names(col) <- paste(names(ann), ann, sep = ".") -->

<!--   ## Create prob colour palette -->

<!--   prob.palette <- colorRamp(c("blue", "yellow", "red")) -->
<!--   prob.palette <- rgb(prob.palette(x$prob)/255) -->

<!--   ## Plot chord diagram -->

<!--   circlize::chordDiagram( -->
<!--     x = dat,  -->
<!--     grid.col = col,  -->
<!--     transparency = 0.5,  -->
<!--     col = "#00000000",  -->
<!--     order = names(col),  -->
<!--     directional = 1,  -->
<!--     direction.type = "arrows",  -->
<!--     annotationTrack = "grid",  -->
<!--     preAllocateTracks = list( -->
<!--       list(track.height = uh(2,'mm')),  -->
<!--       list(track.height = uh(12,'mm')) -->
<!--     ),  -->
<!--     annotationTrackHeight = c(0.01, 0.01), -->
<!--     link.arr.col = prob.palette -->
<!--   ) -->

<!--   ## Plot title -->

<!--   graphics::title(title, cex = 0.6) -->

<!--   ## Plot track region -->

<!--   circlize::circos.trackPlotRegion( -->
<!--     track.index = 2,  -->
<!--     bg.border = 0,  -->
<!--     panel.fun = function(x, y) { -->
<!--       xlim <- get.cell.meta.data("xlim") -->
<!--       ylim <- get.cell.meta.data("ylim") -->
<!--       sector.index <- ann[get.cell.meta.data("sector.numeric.index")] -->
<!--       circos.text( -->
<!--         x = mean(xlim),  -->
<!--         y = mean(ylim),  -->
<!--         sector.index,  -->
<!--         facing = "clockwise",  -->
<!--         niceFacing = TRUE,  -->
<!--         cex = 0.7,  -->
<!--         col = "black" -->
<!--       ) -->
<!--     } -->
<!--   ) -->

<!--   ## Create group colour palette -->

<!--   ids <- unique(names(ann)) -->

<!--   col <- iwanthue(n = length(ids)) -->

<!--   names(col) <- ids -->

<!--   ## Plot highlight sector -->

<!--   for (i in ids) { -->

<!--     df <- ann[names(ann) == i] -->

<!--     circlize::highlight.sector( -->
<!--       sector.index = paste(names(df), df, sep = "."),  -->
<!--       track.index = 1,  -->
<!--       col = col[i],  -->
<!--       lwd = 1, -->
<!--       text = i,  -->
<!--       text.vjust = '0.4cm',  -->
<!--       niceFacing = TRUE -->
<!--     ) -->

<!--   } -->

<!--   # Plot links legend -->

<!--   col <- colorRamp2(c(0, 0.5, 1), c("blue", "yellow", "red")) -->

<!--   lgd <- ComplexHeatmap::Legend(at = c(0, 1), col_fun = col, title = "Prob") -->

<!--   draw(lgd, x = unit(10, "mm"), y = unit(10, "mm"), just = c("left", "bottom")) -->

<!--   # -->

<!--   circos.clear() -->

<!-- } -->
<!-- ``` -->

<!-- ## Source -->

<!-- ```{r} -->
<!-- chordDiagram(x = subset(net, source == 1), title = "Source: 1") -->
<!-- ``` -->

<!-- ## Target -->

<!-- ```{r} -->
<!-- chordDiagram(x = subset(net, source == 1), title = "Source: 1") -->
<!-- ``` -->

<!-- ## Annotation -->

<!-- ```{r} -->
<!-- chordDiagram(x = subset(net, source == 1), title = "Source: 1") -->
<!-- ``` -->

<!-- ## Pathway -->

<!-- ```{r} -->
<!-- chordDiagram(x = subset(net, source == 1), title = "Source: 1") -->
<!-- ``` -->

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
